Efficient pedigree recording for fast population genetics simulation

### 2018-01-13 forwards_paper.tex
Cleaned up the Python implementation of simplify.
Minor updates to the calculations.

The idea of storing genealogical information to speed up simulations is not new.
It was implemented completely in AnA-FiTS~\citep{aberer2013rapid},
but without the critical step of discarding irrelevant genealogical information.
\citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
by keeping track of genealogies over the preceeding 8 generations
and only tracking neutral genotypes for those segments having descendants across this window.
Our approach is similar, but uses genealogies across the entire duration of the simulation.
The embellished pedigree is equivalent to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which has been the subject of substantial study
% under the assumptions of coalescent theory
\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, it is unclear if an ARG-based approach would share
the computational advantages of the data structures we use here \citet{kelleher2016efficient}.

@@

% I may need to cite my student's t-shirt for the following line.
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).  
In this section, we explore the how required computational effort and storage space scale
in a Wright-Fisher model with recombination.

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T = 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
If no coalescent events involve more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$.
Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
and we are measuring tree lengths in units of generations,
the mean number of ancestral recombination events that occur somewhere on the marginal trees
per unit of genetic length.
Any such ancestral recombination might change the marginal gene tree;
if it does, it will add four new edges.
Not all such recombinations actually change the tree topology,
so at most we expect $16N\log(2N)$ new edges after those required to describe the first tree.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
This reduces the initial storage of $120 N^2$ items to
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
the space required after simplification of roughly
\begin{align}
    6N + 32 N \log(2N) .
\end{align}
At $N=10^4$, this reduces the space required to store the history of the entire population
by a factor of 3,700.

>>>>

The idea of storing genealogical information to speed up simulations is not new.
It was implemented in AnA-FiTS~\citep{aberer2013rapid},
but without the critical step of discarding irrelevant genealogical information.
\citet{padhukasahasram2008exploring} obtained impressive speedups for a Wright--Fisher simulation
by keeping track of genealogies over the preceeding 8 generations
and only tracking neutral genotypes for those segments having descendants across this window.
Our approach is similar, but uses genealogies across the entire duration of the simulation.
The embellished pedigree is equivalent to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which has been the subject of substantial study
% under the assumptions of coalescent theory
\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, it is unclear if an ARG-based approach would share
the computational advantages of the data structures we use here~\citep{kelleher2016efficient}.

@@

% I may need to cite my student's t-shirt for the following line.
Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that this method
can dramatically improve simulation performance in practice --
but, how does it perform in theory?
% The simulation results shown in Figures \ref{fig:runtimes_selection} and \ref{fig:relative_speedup_selection} show that
% pedigree tracking greatly speeds up forward-time simulations.  Further, the run times with pedigree tracking become
% near-linear in $\rho$ (Figure \ref{fig:runtimes_selection}).
In this section, we explore how the required computational effort and storage space scale
in a Wright-Fisher model with recombination.

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$NT$ nodes and
$2NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability
($T = 20N$, say, giving a total space requirement of $60N^2$).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $N$ individuals.
If no coalescent events involve more than two lineages, this tree sequence has $2N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $2N\log(N)$.
Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
and we are measuring tree lengths in units of generations,
this is the mean number of ancestral recombination events that occur somewhere on the marginal trees.
Any such ancestral recombination might change the marginal gene tree;
if it does, it will add four new edges.
Not all such recombinations actually change the tree topology,
so at most we expect $8N\log(N)$ new edges after those required to describe the first tree.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
The space complexity after simplification is therefore $O(N + N\log N)$, since
we need $O(N)$ space to store the first marginal tree, and $O(1)$ space
for each of the $O(N \log N)$ trees that follow. Thus, we need $O(N^2)$ space
to store the complete history of the simulation, but only $O(N + N \log N)$
space to store the history that is relevant to the extant population.
